Particle MCMC for Multivariate Stochastic Differential Equations with **`msde`**

### 2021-12-17 tests/dontrun/msde-pmcmc.Rmd
added pkgdown

---
title: "Particle MCMC for Multivariate Stochastic Differential Equations with **`msde`**"
author: "Martin Lysy, Feiyu Zhu"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
bibliography: references.bib
csl: taylor-and-francis-harvard-x.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Particle MCMC with msde}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
\newcommand{\Ep}{\mathrm{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\idm}{\mathrm{I}}
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\N}{\mathcal{N}}

@@

### Posterior sampling via MCMC
We assume only $\Lambda_1$ is unknown (i.e., needs to be estimated). Thus, we introduce a vector `fixed.params` of logical `TRUE` and `FALSE` to indicate if a parameter is fixed or not. Since $\Lambda_1$ is unknown, it is, therefore, not fixed. 
```{r fixed-params}
# only Lambda1 is unknown
fixed.params <- rep(TRUE, bmod$nparams)
names(fixed.params) <- bmod$param.names
fixed.params["Lambda1"] <- FALSE
```
Then we set the prior on the unknown parameter(s) and observations.
```{r prior}
# prior on (Lambda1, Y_0)
hyper <- list(mu = c(0,0), Sigma = diag(2))
names(hyper$mu) <- bmod$data.names
dimnames(hyper$Sigma) <- rep(list(bmod$data.names), 2)
```
We set the total number of posterior samples as `100,000` with the burn-in period as `1,000`. The posterior sampling via MCMC is given as follows:
```{r mcmc, cache = TRUE}
# posterior sampling
nsamples <- 1e5
burn <- 1e3
bpost <- sde.post(bmod, binit, hyper = hyper,
                  fixed.params = fixed.params,
                  nsamples = nsamples, burn = burn)
```
Within `sde.post`, Metropolis-within-Gibbs (MWG) is used. The MWG jump size can be specified as a scalar, a vector or length `nparams + ndims`, or a named vector containing the elements defined by `sde.init$nvar.obs.m[1]` (the missing variables in the first SDE observation) and `fixed.params` (the SDE parameters which are not held fixed). For each MWG random variable, the default jump sizes are `.25 * |initial_value|` when `|initial_value| > 0`, and `1` otherwise. By default, `sde.post` carries out an adaptive MCMC proposal based on @adaptMCMC. At step *n* of the MCMC, the jump size of each MWG random variable is increased or decreased by $\delta(n)$, depending on whether the cumulative acceptance rate is above or below the optimal value of `0.44`.  If $\sigma_n$ is the size of the jump at step `n`, then the next jump size is determined by
$$
\log(\sigma_{n+1}) = \log(\sigma_n) \pm \delta(n), \qquad \delta(n) = \min(.01, 1/n^{1/2}).
$$
For further details of the implementation, please check the description of `sde.post`.

@@

### Analytic sampling via Kalman filter
In order to check the sampling result given by MCMC, we can compare it with the analytic solution given by Kalman filter.
```{r kalman, cache = TRUE}
# analytic posterior
L1.seq <- seq(min(L1.mcmc), max(L1.mcmc), len = 500)
L1.loglik <- sapply(L1.seq, function(l1) {
  lambda <- Lambda0
  lambda[1] <- l1
  mou.loglik(X = YObs, dt = dT, nvar.obs = 1,
             Gamma = Gamma0, Lambda = lambda, Phi = Phi0,
             mu0 = hyper$mu, Sigma0 = hyper$Sigma)
})

@@

### Posterior sampling via particle MCMC
Let's set the number of particles as `100` and just use an adaptive random walk proposal for updating the unknown parameter $\Lambda_1$. Here our aim is to show that even "off-the-shelf" choice can give satisfactory result.
```{r source, echo = FALSE}
source("sde.pmcmc.R")
```
```{r pmcmc, cache = TRUE}
npart <- 100
mwg.sd <- 1
delta <- .618
ppost <- sde.pmcmc(model = bmod, binit, theta0, fixed.params, hyper,
                   nsamples, npart, dT,
                   resample = "multi", threshold = 0.5, mwg.sd, delta)
```
In general we have the following proposal:
$$
\theta_{j, prop} \sim \N(\theta_{j, curr}, \sigma^2_{n,j})
$$
where $\theta_{j,prop}$ is the proposed (new) $j$-th parameter and $\theta_{j,curr}$ is the current (or old) $j$-th parameter. Besides, the adaptive standard deviation $\sigma_{n+1,j}$ at the $n+1$-th iteration for the $j$-th parameter is determined by
$$
\sigma_{n+1,j} = \exp\left(\log(\sigma_{n,j}) \pm \frac{\delta}{n} \right).
$$
Based on the above adaptive rule, we increase/decrease $\sigma_{n,j}$ at each step if the previous draw was accepted/rejected. The amount of adaptation will goes to $0$ as the number $n$ of iterations goes large. This kind of adaptation can preserve the MCMC stationary distribution.

@@

Finally, we extract the posterior samples of $\Lambda_1$ which will be used for comparison in the following subsection.
```{r}
# posterior mean of theta
L1.pmcmc <- ppost$params[ ,!fixed.params]
```
### Result comparison 
Similarly, we compare the histogram of posterior samples of $\Lambda_1$ given by particle MCMC with the analytic curve solved by Kalman filter.
```{r pmcmc-plot, fig.width = 10, fig.height = 5, out.width = "90%"}
# compare particle MCMC with Kalman filter
hist(L1.pmcmc, breaks = 100, freq = FALSE,
     main = expression(p(Lambda[1]*" | "*bold(Y)[1])),
     xlab = expression(Lambda[1]))
lines(L1.seq, L1.Kalman, col = "red")
legend("topright", legend = c("Analytic", "PMCMC"),
       pch = c(NA, 22), lty = c(1, NA), col = c("red", "black"))
```

>>>>

---
title: "Particle MCMC for Multivariate Stochastic Differential Equations with **`msde`**"
author: "`r readLines('authors.txt')`"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
bibliography: references.bib
csl: taylor-and-francis-harvard-x.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Particle MCMC with msde}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
\newcommand{\Ep}{\mathrm{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\idm}{\mathrm{I}}
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\N}{\mathcal{N}}

@@

### Posterior sampling via MCMC
We assume only $\Lambda_1$ is unknown (i.e., needs to be estimated). Thus, we introduce a vector `fixed.params` of logical `TRUE` and `FALSE` to indicate if a parameter is fixed or not. Since $\Lambda_1$ is unknown, it is, therefore, not fixed. 
```{r fixed-params}
# only Lambda1 is unknown
fixed.params <- rep(TRUE, bmod$nparams)
names(fixed.params) <- bmod$param.names
fixed.params["Lambda1"] <- FALSE
```
Then we set the prior on the unknown parameter $\Lambda_1$ and the latent observation $Y_2$. The other elements are given a Lebesgue prior.
$$
\Lambda_1, Y_2 \overset{iid}{\sim} \N(0,1), \quad \pi(Y_1, \Gamma, \Lambda_2, \Phi | \Lambda_1, Y_2) \propto 1.
$$ 
```{r prior}
# prior on (Lambda1, Y2)
hyper <- list(mu = c(0,0), Sigma = diag(2))
names(hyper$mu) <- c("Lambda1", "Y2")
dimnames(hyper$Sigma) <- rep(list(c("Lambda1", "Y2")), 2)
```
We set the total number of posterior samples as `100,000` with the burn-in period as `1,000`. The posterior sampling via MCMC is given as follows:
```{r mcmc}
# posterior sampling
nsamples <- 1e5
burn <- 1e3
bpost <- sde.post(bmod, binit, hyper = hyper,
                  fixed.params = fixed.params,
                  nsamples = nsamples, burn = burn)
```
Within `sde.post`, Metropolis-within-Gibbs (MWG) is used. The MWG jump size can be specified as a scalar, a vector or length `nparams + ndims`, or a named vector containing the elements defined by `sde.init$nvar.obs.m[1]` (the missing variables in the first SDE observation) and `fixed.params` (the SDE parameters which are not held fixed). For each MWG random variable, the default jump sizes are `.25 * |initial_value|` when `|initial_value| > 0`, and `1` otherwise. By default, `sde.post` carries out an adaptive MCMC proposal based on @adaptMCMC. At step *n* of the MCMC, the jump size of each MWG random variable is increased or decreased by $\delta(n)$, depending on whether the cumulative acceptance rate is above or below the optimal value of `0.44`.  If $\sigma_n$ is the size of the jump at step `n`, then the next jump size is determined by
$$
\log(\sigma_{n+1}) = \log(\sigma_n) \pm \delta(n), \qquad \delta(n) = \min(.01, 1/n^{1/2}).
$$
For further details of the implementation, please check the description of `sde.post`.

@@

### Analytic sampling via Kalman filter
In order to check the sampling result given by MCMC, we can compare it with the analytic solution given by Kalman filter.
```{r kalman, cache = TRUE}
# analytic posterior
L1.seq <- seq(min(L1.mcmc), max(L1.mcmc), len = 500)
L1.loglik <- sapply(L1.seq, function(l1) {
  lambda <- Lambda0
  lambda[1] <- l1
  tmp <- mou.loglik(X = YObs, dt = dT, nvar.obs = 1,
                  Gamma = Gamma0, Lambda = lambda, Phi = Phi0,
                  mu0 = c(0, 0), Sigma0 = diag(2))
  tmp <- tmp + dnorm(l1, 0, 1, log = TRUE) # add the prior log density of Lambda1
  return(tmp)
})

@@

### Posterior sampling via particle MCMC
Let's set the number of particles as `100` and just use an adaptive random walk proposal for updating the unknown parameter $\Lambda_1$. Here our aim is to show that even "off-the-shelf" choice can give satisfactory result.
```{r source, echo = FALSE}
source(system.file("proj", "sde.pmcmc.R", package = "msde"))
```
```{r pmcmc, cache = TRUE}
npart <- 100
rw.sd <- 1
delta <- .7
ppost <- sde.pmcmc(model = bmod, binit, theta0, fixed.params, hyper,
                   nsamples, npart, dT,
                   resample = "multi", threshold = 0.5, rw.sd, delta)
```
In general we have the following proposal:
$$
\theta_{j, prop} \sim \N(\theta_{j, curr}, \sigma^2_{n,j})
$$
where $\theta_{j,prop}$ is the proposed (new) $j$-th parameter and $\theta_{j,curr}$ is the current (or old) $j$-th parameter. Besides, the adaptive standard deviation $\sigma_{n+1,j}$ at the $n+1$-th iteration for the $j$-th parameter is determined by
$$
\sigma_{n+1,j} = \exp\left(\log(\sigma_{n,j}) \pm \frac{\delta}{n} \right).
$$
Based on the above adaptive rule, we increase/decrease $\sigma_{n,j}$ at each step if the previous draw was accepted/rejected. The amount of adaptation will goes to $0$ as the number $n$ of iterations goes large. This kind of adaptation can preserve the MCMC stationary distribution.

@@

Finally, we extract the posterior samples of $\Lambda_1$ which will be used for comparison in the following subsection.
```{r}
# posterior samples
L1.pmcmc <- ppost$params[ ,!fixed.params]
```
### Result comparison 
Similarly, we compare the histogram of posterior samples of $\Lambda_1$ given by particle MCMC with the analytic curve solved by Kalman filter.
```{r pmcmc-plot, fig.width = 10, fig.height = 5, out.width = "90%"}
# compare particle MCMC with Kalman filter
hist(L1.pmcmc, breaks = 100, freq = FALSE,
     main = expression(p(Lambda[1]*" | "*bold(Y)[1])),
     xlab = expression(Lambda[1]))
lines(L1.seq, L1.Kalman, col = "red")
legend("topright", legend = c("Analytic", "PMCMC"),
       pch = c(NA, 22), lty = c(1, NA), col = c("red", "black"))
```
